Ethnicity Sensitive Author Disambiguation Using Semi-supervised Learning

### 2016-06-27 paper/author_disambiguation.tex
Modify the affiliation line of Hussein
june 1 deadline
Signed-off-by: Mateusz Susik <mateuszsusik@gmail.com>

\setlength{\intextsep}{5pt plus 0pt minus 0pt}
\def\baselinestretch{0.99}

\newcommand{\longpage}{\enlargethispage{\baselineskip}}
\newcommand{\verylongpage}{\enlargethispage{2\baselineskip}}

@@

% paper title
% can use linebreaks \\ within to get better formatting as desired
\title{Ethnicity sensitive author disambiguation using semi-supervised learning}

@@

\institute{CERN, Switzerland,\\
\email{\{g.louppe, eamonn.james.maguire\}@cern.ch},
\and
ISH-LYON (USR 3385), CNRS, France,\\
\email{hussein.al-natsheh@ish-lyon.cnrs.fr}
\and
University of Warsaw, Poland,\\
\email{msusik@student.uw.edu.pl}
}

@@

Author name disambiguation in bibliographic databases is the problem of grouping
together scientific publications written by the same person, accounting for
potential homonyms and/or synonyms. Among solutions to this problem, digital
libraries are increasingly offering tools for authors to manually curate their
publications and claim those that are theirs. Indirectly, these tools allow for
the inexpensive collection of large annotated training data, which can be
further leveraged to build a complementary automated disambiguation system
capable of inferring patterns for identifying publications written by the same
person. Building on more than 1 million crowdsourced
annotations that we publicly release, we propose an automated author disambiguation solution exploiting
this data (i) to learn an accurate classifier for identifying coreferring
authors and (ii) to guide the clustering of scientific publications by distinct
authors in a semi-supervised way. To the best of our knowledge, our analysis is
the first to be carried out on data of this size and coverage. With respect to
the state of the art, we validate the general pipeline used in most existing
solutions, and improve by: (i) proposing phonetic-based blocking strategies,
thereby increasing recall; (ii) adding strong ethnicity-sensitive features for
learning a linkage function, thereby tailoring disambiguation to non-Western
author names whenever necessary; and (iii) showing the importance of balancing
negative and positive examples when learning the linkage function.

@@

In the case of multiple surnames (cases 2 and 3), we propose to block
signatures in two phases.
In the first phase, all the signatures with a single surname are clustered together.
Every different surname token creates a new block.
In the second phase, the signatures with multiple surnames are compared
with the blocks for the first and last surname.
If the first surnames of an author were already used as the last given names on some of the signatures, the
new signature is assigned to the block of the last surname (case 2).
Otherwise, the signature is assigned to the block of the first surname (case 3).
Finally, to prevent the creation of too large blocks, signatures are further divided
along their first given name initial.
Cases 4 and 5 are not explicitly handled.

@@

\subsubsection{Input features.} Following previous works, pairs of signatures $(s_1, s_2)$ are first transformed to vectors $v \in \mathbb{R}^p$
by building so-called similarity profiles \cite{treeratpituk2009disambiguating} on which supervised learning is carried out.
In this work, we design and evaluate fifteen standard input
features based on the comparison of signature fields, as reported in the first
half of Table~\ref{table:features}.
As an illustrative example, the \textit{Full name} feature corresponds to the similarity between the (full)
author name fields of the two signatures, as measured using as combination
operator the cosine similarity between their respective $(n,m)$-\emph{TF-IDF} vector
representations\footnote{$(n,m)-$\emph{TF-IDF} vectors are \emph{TF-IDF} vectors computed
from $n$, $n+1$, ..., $m$-grams.}.

@@

All the solutions proposed in this work are evaluated on data extracted from
the \emph{INSPIRE} portal \cite{gentil2009information}, a digital library for scientific literature in
high-energy physics.
Overall, the portal holds more than 1 million publications ${\cal P}$,
forming in total a set ${\cal S}$ of more than 10 million signatures.
Out of these, around 13\% have been \textit{claimed} by their
original authors, marked as such by professional curators or automatically assigned to their true authors thanks
to persistent identifiers provided by publishers or other sources.
Together, they constitute a trusted set $({\cal S}^\prime, {\cal C}^\prime)$ of 15388 distinct individuals sharing
36340 unique author names spread within 1201763 signatures on 360066
publications. This data covers several decades in time and dozens of author
nationalities worldwide.

@@

\begin{small}
\begin{table*}
\caption{Average precision, recall and f-measure scores on test folds. - components correspond to the state-of-the-art choices.}
\label{table:results}
\centering
\begin{tabular}{|l|c c c | c c c|}
  \hline
                       & \multicolumn{3}{|c|}{\textbf{B3}} & \multicolumn{3}{|c|}{\textbf{Pairwise}}\\
  \textbf{Description} & $P$ & $R$ & $F$ & $P$ & $R$ & $F$ \\
  \hline
  \hline
Baseline & 0.9024 & 0.9828 & 0.9409  & 0.8298 & 0.9776 & 0.8977 \\
\hline
\underline{Blocking = SFI} & 0.9901 & 0.9760 & 0.9830  & 0.9948 & 0.9738 & 0.9842\\
Blocking = Double metaphone & 0.9856 & 0.9827 & 0.9841  & 0.9927 & 0.9817 & 0.9871 \\
Blocking = NYSIIS & 0.9875 & 0.9826 & \textbf{0.9850}  & 0.9936 & 0.9814 & \textbf{0.9875} \\
Blocking = Soundex & 0.9886 & 0.9745 & 0.9815 & 0.9935 & 0.9725 & 0.9828 \\
\hline
\underline{Classifier = GBRT} & 0.9901 & 0.9760 & 0.9830   & 0.9948 & 0.9738 & 0.9842 \\
Classifier = Random Forests & 0.9909 & 0.9783 & \textbf{0.9846}  & 0.9957 & 0.9752 & \textbf{0.9854} \\
Classifier = Linear Regression & 0.9749 & 0.9584 & 0.9666 & 0.9717 & 0.9569 & 0.9643 \\
\hline
Training pairs = Non-blocked, uniform & 0.9793 & 0.9630 & 0.9711  & 0.9756 & 0.9629 & 0.9692 \\
Training pairs = Blocked, uniform & 0.9854 & 0.9720 & 0.9786  & 0.9850 & 0.9707 & 0.9778 \\
\underline{Training pairs = Blocked, balanced} & 0.9901 & 0.9760 & \textbf{0.9830}  & 0.9948 & 0.9738 & \textbf{0.9842} \\
\hline
\underline{Clustering = Average linkage} & 0.9901 & 0.9760 & \textbf{0.9830}  & 0.9948 & 0.9738 & \textbf{0.9842} \\
Clustering = Single linkage & 0.9741 & 0.9603 & 0.9671  & 0.9543 & 0.9626 & 0.9584 \\
Clustering = Complete linkage & 0.9862 & 0.9709 & 0.9785  & 0.9920 & 0.9688 & 0.9803 \\
\hline
No cut (baseline) & 0.9024 & 0.9828 & 0.9409  & 0.8298 & 0.9776 & 0.8977 \\
Global cut & 0.9892 & 0.9737 & 0.9814  & 0.9940 & 0.9727 & 0.9832 \\
\underline{Block cut} & 0.9901 & 0.9760 & \textbf{0.9830}  & 0.9948 & 0.9738 & \textbf{0.9842} \\
\hline
Combined best settings & 0.9888 & 0.9848 & \textbf{0.9868}  & 0.9951 & 0.9831 & \textbf{0.9890} \\
Best settings without ethnicity features & 0.9862 & 0.9819 & 0.9841 & 0.9937 & 0.9815 & 0.9876 \\
  \hline
\end{tabular}
\end{table*}
\end{small}

@@

The good precision of the state-of-the-art ($0.9901$), but its
lower recall ($0.9760$) suggest that the blocking strategy might be the
limiting factor to further overall improvements.
Our experiments showed the maximum $B3$ recall (\ie, if within a block, all signatures were clustered optimally) for SFI is $0.9828$, which corroborates the
estimation of this technique on real data by \cite{torvik2009author}.
At the price of fewer and therefore slightly larger blocks, the
proposed phonetic-based blocking strategies show better maximum recall (all
around $0.9905$), thereby pushing further the upper bound on the maximum
performance of author disambiguation.
Let us remind however that the reported maximum recalls for the blocking strategies using phonetization are
also raised due to the better handling of multiple surnames, as described in Section \ref{methods:blocking}.

As Table~\ref{table:results} shows, switching to either Double metaphone or NYSIIS phonetic-based blocking allows
to improve the overall F-measure score.
In particular, the NYSIIS-based phonetic blocking shows to be the most effective when applied
to the state-of-the-art (with an F-measure of $0.9850$) while also being the most
efficient computationally (with 10857 blocks versus 12978 for the baseline).

@@

Second, properly constructing a training set of positive and negative pairs of
signatures from which to learn a linkage function yields a significant
improvement.
A random sampling of positive and negative pairs, without taking
blocking into account, significantly impacts the overall performance
($0.9711$). When pairs are drawn only from blocks, performance increases
($0.9786$), which confirms our intuition that $d$ should be built only from
pairs it will be used to eventually cluster. Finally, making the classification
problem more difficult by oversampling complex cases proves to be relevant,
by further improving the disambiguation results ($0.9830$).

>>>>

\setlength{\intextsep}{5pt plus 0pt minus 0pt}
% \def\baselinestretch{0.99}

\newcommand{\longpage}{\enlargethispage{\baselineskip}}

@@

% paper title
% can use linebreaks \\ within to get better formatting as desired
\title{Ethnicity Sensitive Author Disambiguation Using Semi-supervised Learning}

@@

\institute{CERN, Switzerland,\\
\email{\{g.louppe, eamonn.james.maguire\}@cern.ch},
\and
Univ Lyon, ISH Lyon, USR 3385, CNRS, and ERIC lab., EA 3083, Lyon 2, France,\\
\email{hussein.al-natsheh@ish-lyon.cnrs.fr}
\and
University of Warsaw, Poland,\\
\email{msusik@student.uw.edu.pl}
}

@@

Building on more than 1 million crowdsourced
annotations that we publicly release, we propose a new automated disambiguation
solution exploiting
this data (i) to learn an accurate classifier for identifying coreferring
authors and (ii) to guide the clustering of scientific publications by distinct
authors in a semi-supervised way. To the best of our knowledge, our analysis is
the first to be carried out on data of this size and coverage. With respect to
the state of the art, we validate the general pipeline used in most existing
solutions, and improve by: (i) proposing new phonetic-based blocking strategies,
thereby increasing recall; (ii) adding strong ethnicity-sensitive features for
learning a linkage function, thereby tailoring disambiguation to non-Western
author names whenever necessary; and (iii) showing the importance of balancing
negative and positive examples when learning the linkage function.

@@

In the case of multiple surnames (cases 2 and 3), we propose to block
signatures in two phases.
In the first phase, all the signatures with a single surname are clustered together.
Every different surname token creates a new block.
In the second phase, the signatures with multiple surnames are compared
with the blocks for the first and last surname.
If the first surnames of an author were already used as the last given names on some of the signatures, the
new signature is assigned to the block of the last surname (case 2).
Otherwise, the signature is assigned to the block of the first surname (case 3).
Finally, to prevent the creation of too large blocks, signatures are further divided
along their first given name initial.
The biggest limitation of this method is leaving the cases 4 and 5 unhandled.
This method might result in blocking together signatures from many different authors
with similar names.

@@

\subsubsection{Input features.} Following previous works, pairs of signatures $(s_1, s_2)$ are first transformed to vectors $v \in \mathbb{R}^p$
by building so-called similarity profiles \cite{treeratpituk2009disambiguating} on which supervised learning is carried out.
In this work, we design and evaluate fifteen standard input
features \cite{ferreira2012brief, levin2012citation} based on the comparison of signature fields, as reported in the first
half of Table~\ref{table:features}.
As an illustrative example, the \textit{Full name} feature corresponds to the similarity between the (full)
author name fields of the two signatures, as measured using as combination
operator the cosine similarity between their respective $(n,m)$-\emph{TF-IDF} vector
representations\footnote{$(n,m)-$\emph{TF-IDF} vectors are \emph{TF-IDF} vectors computed
from $n$, $n+1$, ..., $m$-grams.}.

@@

All the solutions proposed in this work are evaluated on data extracted from
the \emph{INSPIRE} portal \cite{gentil2009information}, a digital library for scientific literature in
high-energy physics.
Overall, the portal holds more than 1 million publications ${\cal P}$,
forming in total a set ${\cal S}$ of more than 10 million signatures.
Out of these, around 13\% have been \textit{claimed} by their
original authors, marked as such by professional curators or automatically assigned to their true authors thanks
to persistent identifiers provided by publishers or other sources.
Together, they constitute a trusted set $({\cal S}^\prime, {\cal C}^\prime)$ of 15,388 distinct individuals sharing
36,340 unique author names spread within 1,201,763 signatures on 360,066
publications. This data covers several decades in time and dozens of author
nationalities worldwide.

@@

\begin{small}
\begin{table*}
\caption{Average precision, recall and f-measure scores on test folds. Components correspond to the state-of-the-art choices.}
\label{table:results}
\centering
\begin{tabular}{|l|c c c | c c c|}
  \hline
                       & \multicolumn{3}{|c|}{\textbf{B3}} & \multicolumn{3}{|c|}{\textbf{Pairwise}}\\
  \textbf{Description} & $P$ & $R$ & $F$ & $P$ & $R$ & $F$ \\
  \hline
  \hline
Baseline & 0.9024 & 0.9828 & 0.9409  & 0.8298 & 0.9776 & 0.8977 \\
\hline
\underline{Blocking = SFI} & 0.9901 & 0.9760 & 0.9830  & 0.9948 & 0.9738 & 0.9842\\
Blocking = Double metaphone & 0.9856 & 0.9827 & 0.9841  & 0.9927 & 0.9817 & 0.9871 \\
Blocking = NYSIIS & 0.9875 & 0.9826 & \textbf{0.9850}  & 0.9936 & 0.9814 & \textbf{0.9875} \\
Blocking = Soundex & 0.9886 & 0.9745 & 0.9815 & 0.9935 & 0.9725 & 0.9828 \\
\hline
\underline{Classifier = GBRT} & 0.9901 & 0.9760 & 0.9830   & 0.9948 & 0.9738 & 0.9842 \\
Classifier = Random Forests & 0.9909 & 0.9783 & \textbf{0.9846}  & 0.9957 & 0.9752 & \textbf{0.9854} \\
Classifier = Linear Regression & 0.9749 & 0.9584 & 0.9666 & 0.9717 & 0.9569 & 0.9643 \\
\hline
Training pairs = Non-blocked, uniform & 0.9793 & 0.9630 & 0.9711  & 0.9756 & 0.9629 & 0.9692 \\
Training pairs = Blocked, uniform & 0.9854 & 0.9720 & 0.9786  & 0.9850 & 0.9707 & 0.9778 \\
\underline{Training pairs = Blocked, balanced} & 0.9901 & 0.9760 & \textbf{0.9830}  & 0.9948 & 0.9738 & \textbf{0.9842} \\
\hline
\underline{Clustering = Average linkage} & 0.9901 & 0.9760 & \textbf{0.9830}  & 0.9948 & 0.9738 & \textbf{0.9842} \\
Clustering = Single linkage & 0.9741 & 0.9603 & 0.9671  & 0.9543 & 0.9626 & 0.9584 \\
Clustering = Complete linkage & 0.9862 & 0.9709 & 0.9785  & 0.9920 & 0.9688 & 0.9803 \\
\hline
No cut (baseline) & 0.9024 & 0.9828 & 0.9409  & 0.8298 & 0.9776 & 0.8977 \\
Global cut & 0.9892 & 0.9737 & 0.9814  & 0.9940 & 0.9727 & 0.9832 \\
\underline{Block cut} & 0.9901 & 0.9760 & \textbf{0.9830}  & 0.9948 & 0.9738 & \textbf{0.9842} \\
\hline
Combined best settings & 0.9888 & 0.9848 & \textbf{0.9868}  & 0.9951 & 0.9831 & \textbf{0.9890} \\
Best settings without ethnicity features & 0.9862 & 0.9819 & 0.9841 & 0.9937 & 0.9815 & 0.9876 \\
  \hline
\end{tabular}
\end{table*}
\end{small}

@@

The good precision of the state-of-the-art ($0.9901$), but its
lower recall ($0.9760$) suggest that the blocking strategy might be the
limiting factor to further overall improvements.
Our experiments showed the maximum $B3$ recall (\ie, if within a block, all signatures were clustered optimally) for SFI is $0.9828$, which corroborates the
estimation of this technique on real data by \cite{torvik2009author}.
At the price of fewer and therefore slightly larger blocks, the
proposed phonetic-based blocking strategies show better maximum recall (all
around $0.9905$). Better recall pushes further the upper bound on the maximum
performance of author disambiguation, as the signatures that belong to the same author
and different groups can not be clustered together by our algorithm.
Let us remind that the reported maximum recalls for the blocking strategies using phonetization are
also raised due to the better handling of multiple surnames, as described in Section \ref{methods:blocking}.

As Table~\ref{table:results} shows, switching to either Double metaphone or NYSIIS phonetic-based blocking allows
to improve the overall F-measure score.
In particular, the NYSIIS-based phonetic blocking shows to be the most effective when applied
to the state-of-the-art (with an F-measure of $0.9850$) while also being the most
efficient computationally (with 10,857 blocks versus 12,978 for the baseline).

@@

Second, properly constructing a training set of positive and negative pairs of
signatures from which to learn a linkage function yields a significant
improvement.
A random sampling of positive and negative pairs, without taking
blocking into account, significantly impacts the overall performance
($0.9711$). When pairs are drawn only from blocks, performance increases
($0.9786$), which confirms our intuition that $d$ should be built only from
pairs it will be used to eventually cluster. Finally, making the classification
problem more difficult by oversampling complex cases (see \ref{methods:linkage}) proves to be relevant,
by further improving the disambiguation results ($0.9830$).
